SU($n$) ferromagnets near the Heisenberg critical point

### 2020-04-03 collective_sun/sun_perturbation.tex
finish first draft of introduction

\title{SU($n$) spin models near the Heisenberg critical point}%
\author{Michael A. Perlin}%
\date{\today}

@@

SU($n$) symmetries play an important role in contemporary physics.
Underpinning much of high energy physics, the celebrated SU($n$) gauge
theory known as Yang-Mills theory is central to our understanding of
the electroweak and strong forces.  Extensions of Yang-Mills and
SU($n$) symmetry feature in the most well-studied examples of
holographic duality\cite{maldacena1999largen} and the connection
between entanglement and gravity\cite{ryu2006holographic} through the
anti-de Sitter / conformal field theory (AdS/CFT) correspondence.  In
a condensed matter setting, SU(2) appears ubiquitously as a symmetry
of the Hubbard model, with important consequences for the study of
quantum magnetism and high temperature
superconductivity\cite{lee2006doping}.  The extension of the SU(2)
Hubbard model to SU($n$) has led to predictions of exotic phases of
matter such as chiral spin liquids\cite{hermele2009mott,
  hermele2011topological, chen2016syntheticgaugefield,
  nataf2016chiral}, as well as the potential to perform universal
topological quantum computation\cite{freedman2004class,
  nayak2008nonabelian}.  Furthermore, the consideration of disordered
SU($n$) spin models has opened analytically tractable avenues for
studying quantum chaos and information
scrambling\cite{sachdev1993gapless, bentsen2019integrable}.

Last but not least, SU($n$) symmetries appear naturally in the physics
of atomic, molecular, and optical (AMO)
systems\cite{gorshkov2010twoorbital, beverland2016realizing,
  cazalilla2014ultracold, taie2012su, hofrichter2016direct,
  cappellini2014direct, scazza2014observation, zhang2014spectroscopic,
  goban2018emergence, perlin2019effective} through the independence of
atomic orbital and interaction parameters on the $n$ nuclear spin
states of individual atoms, with e.g.~$n=10$ in the case of
${}^{87}$Sr.  As a result, experiments can directly probe the SU($n$)
Hubbard model, leading to experimental observations of SU($n$) Hubbard
phases and phase transitions\cite{taie2012su, hofrichter2016direct},
two-orbital SU($n$) magnetism\cite{cappellini2014direct,
  scazza2014observation, zhang2014spectroscopic}, and multi-body
SU($n$)-symmetric interactions\cite{goban2018emergence,
  perlin2019effective}.  The tremendous theoretical significance of
SU($n$) symmetries makes it all the more exciting that they appear in
experimental platforms with exquisite degrees of microscopic control,
thereby enabling the detailed study of these symmetries' consequences
for fundamental questions in physics, as well as their use in
practical technological applications.  SU(2)-symmetric interactions,
for example, can play an important role in the development of quantum
sensors that make use of entanglement to surpass classical limits on
measurement precision\cite{he2019engineering}\red{[cite XXZ squeezing
  paper]}.  The exciting prospect of similarly exploiting larger
SU($n$) symmetries to achieve a technological advantage is still an
unexplored avenue with enormous potential.

\red{[finish introduction]}

\vfill

\vspace{3cm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The SU($n$) Heisenberg model}

@@

We consider an array of $N$ multilevel spins with SU($n$)-symmetric
interactions that can be written in the form
\begin{align}
  H_0 = \sum_{\set{p,q}\in\C_N\p{2}} h_{pq} \Pi_{pq},
  &&
  \Pi_{pq} \equiv \sum_{\mu,\nu\in\ZZ_n} s_{\mu\nu}^{(p)} s_{\nu\mu}^{(q)},
  \label{eq:H_0}
\end{align}
where $p,q$ index individual spins; $\C_N\p{2}$ is the set of all
subsets (``choices'') of two spins out of $N$; $h_{pq}$ are scalar
coupling constants; $\mu,\nu$ index states in an orthonormal basis for
the $n$-dimensional Hilbert space of a single spin; $\ZZ_n$ is the set
of integers modulo $n$; the operator
$s_{\mu\nu}^{(p)}\equiv\op{\mu}{\nu}_p$ flips the state of spin $p$ to
$\ket\nu$ from $\ket\mu$; and the operator $\Pi_{pq}$ permutes the
states of spins $p$ and $q$.  The permutation operator $\Pi_{pq}$ is
essentially a multilevel generalization of the SU(2)-symmetric spin
interaction $\v s_p\c\v s_q$, with $\v s\equiv\p{\v s_\x,s_\y,s_\z}$,
$s_\alpha\equiv\sigma_\alpha/2$, and $\sigma_\alpha$ a single-spin
Pauli operator.

@@

The PS manifold $\M_0$ is spanned by a basis of states
$\ket{m}=\ket{m_1,m_2,\cdots,m_n}$ with a definite occupation number
$m_\mu$ of the single-spin state $\mu$, and $\sum_\mu m_\mu=N$.  We
denote the set of all such assignments of $N$ (identical) spins to $n$
(distinct) states by $\A_n\p{N}$, such that our standard basis for
$\M_0$ is $\set{\ket{m}:m\in\A_n\p{N}}$.  The dimension of $\M_0$ is
then the number of such assignments, which is straightforward to
determine with a standard ``stars and bars'' calculation:
$\abs{\A_n\p{N}} = {N+n-1 \choose n-1} \sim N^{n-1}$.  In the case of
$n=2$, for example, the PS manifold $\M_0$ is precisely the Dicke
manifold\cite{dicke1954coherence} of $N$ qubits, spanned by
${N+2-1\choose2-1}=N+1$ Dicke states $\ket{m_0,m_1}$ for which
$m_0+m_1=N$ and the integer $m_0$ ($m_1$) indicates the total
occupation number of qubit state $\ket{0}$ ($\ket{1}$).  The energy of
any PS state $\ket\psi\in\M_0$ with respect to the interaction
Hamiltonian $H_0$ in \eqref{eq:H_0} is
\begin{align}
  E_0 = \sum_{\set{p,q}\in\C_N\p{2}} h_{pq}.
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Breaking SU($n$) symmetry}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Beyond ground-state perturbation theory}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The XXZ model}

>>>>

\title{SU($n$) ferromagnets near the Heisenberg critical point}%
\author{Michael A. Perlin}%
\date{\today}

@@

SU($n$) symmetries play an important role in physics.  Underpinning
much of high energy physics, the celebrated SU($n$) gauge theory known
as Yang-Mills theory is central to our understanding of the
electroweak and strong forces.  Extensions of Yang-Mills and SU($n$)
symmetry feature in the most well-studied examples of holographic
duality\cite{maldacena1999largen} and the connection between
entanglement and gravity\cite{ryu2006holographic} through the anti-de
Sitter / conformal field theory (AdS/CFT) correspondence.  In a
condensed matter setting, SU(2) appears ubiquitously as a symmetry of
the Hubbard model, with important consequences for the study of
quantum magnetism and high temperature
superconductivity\cite{lee2006doping}.  The extension of the SU(2)
Hubbard model to SU($n$) has led to predictions of exotic phases of
matter such as chiral spin liquids\cite{hermele2009mott,
  hermele2011topological, chen2016syntheticgaugefield,
  nataf2016chiral}, as well as the potential to perform universal
topological quantum computation\cite{freedman2004class,
  nayak2008nonabelian}.  Furthermore, the consideration of disordered
SU($n$) spin models has opened analytically tractable avenues for
studying quantum chaos and information
scrambling\cite{sachdev1993gapless, bentsen2019integrable}.

Last but not least, SU($n$) symmetries appear naturally in the physics
of atomic, molecular, and optical (AMO)
systems\cite{gorshkov2010twoorbital, beverland2016realizing,
  cazalilla2014ultracold, taie2012su, hofrichter2016direct,
  cappellini2014direct, scazza2014observation, zhang2014spectroscopic,
  goban2018emergence, perlin2019effective} through the independence of
atomic orbital and interaction parameters on the $n$ nuclear spin
states of individual atoms, with e.g.~$n=10$ in the case of
${}^{87}$Sr.  As a result, experiments can directly probe the SU($n$)
Hubbard model, leading to experimental observations of SU($n$) Hubbard
phases and phase transitions\cite{taie2012su, hofrichter2016direct},
two-orbital SU($n$) magnetism\cite{cappellini2014direct,
  scazza2014observation, zhang2014spectroscopic}, and multi-body
SU($n$)-symmetric interactions\cite{goban2018emergence,
  perlin2019effective}.  The tremendous theoretical significance of
SU($n$) symmetries makes it all the more exciting that they appear in
experimental platforms with exquisite degrees of microscopic control,
thereby enabling the detailed study of these symmetries' consequences
for fundamental questions in physics, as well as their practical use
in technological applications.  SU(2)-symmetric interactions, for
example, can play an important role in the development of quantum
sensors that make use of entanglement to surpass classical limits on
measurement precision\cite{he2019engineering}\red{[cite XXZ squeezing
  paper]}.  The exciting prospect of similarly exploiting larger
SU($n$) symmetries to achieve a technological advantage is still an
unexplored avenue of research with enormous potential.

In this work, we generalize some of the techniques used to harness
SU(2)-symmetric spin interactions for metrological advantage to the
case of SU($n$).  In doing so, we develop a low-energy effective
theory of gapped SU($n$) ferromagnets with weakly broken SU($n$)
symmetry, e.g.~by the presence of external fields or additional
interaction terms.  The restriction of our theory to ferromagnetic
interactions can be relaxed to the assumption of
\begin{enumerate*}
\item an initial state in the manifold of permutationally-symmetric
  states, a multilevel generalization of the Dicke manifold for
  two-level spins, together with the assumption that
\item this manifold is gapped away from states that break
  permutational symmetry.
\end{enumerate*}
In addition to working out the effective Hamiltonian induced by weak
SU($n$) symmetry breaking through second order in perturbation theory,
we develop a diagrammatic method to systematically compute all matrix
elements of arbitrary spin operators restricted manifolds of weakly
broken permutation symmetry, thereby enabling the numerical simulation
of dynamics induced by larger, non-perturbative modifications to the
SU($n$) Heisenberg model.  While the latter simulations generally
accumulate uncontrolled truncation errors, we identify criteria under
which numerical results correctly capture the qualitative dynamical
behavior of a system.

In the following section, \ref{sec:bare_sun}, we introduce the SU($n$)
Heisenberg model, finding a convenient way to express its Hamiltonian
in terms of permutation operators that are agnostic to the value of
$n$.  In Section \ref{sec:breaking_sun}, we consider the introduction
of general spin operators, such as inhomogeneous external fields, that
weakly break SU($n$) symmetry.  We sketch out a derivation of the
effective Hamiltonian induced by these terms through second order in
perturbation theory, deferring more detailed calculations to
Appendices \ref{sec:eigenstates} and \ref{sec:operator_product}.
Finally, we discuss a numerical method to simulate dynamics within
manifolds of weak permutational symmetry breaking in Section
\ref{sec:shell_model}, before concluding in Section
\ref{sec:conclusions}.  Wherever possible, we will fall back to the
broadly familiar case of SU(2) in order to facilitate an understanding
of the general result for SU($n$).

\red{[re-visit / revise the introduction after finishing the remaining
  text]}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The SU($n$) Heisenberg model}
\label{sec:bare_sun}

@@

We consider an array of $N$ multilevel spins with SU($n$)-symmetric
interactions that can be written in the form
\begin{align}
  H_0 = \sum_{\set{p,q}\in\C_N\p{2}} h_{pq} \Pi_{pq},
  &&
  \Pi_{pq} \equiv \sum_{\mu,\nu\in\ZZ_n} s_{\mu\nu}^{(p)} s_{\nu\mu}^{(q)},
  \label{eq:H_0}
\end{align}
where $p,q$ index individual spins; $\C_N\p{2}$ is the set of all
subsets (``choices'') of two spins out of $N$; $h_{pq}$ are scalar
coupling constants; $\mu,\nu$ index states in an orthonormal basis for
the $n$-dimensional Hilbert space of a single spin; $\ZZ_n$ is the set
of integers modulo $n$; the operator
$s_{\mu\nu}^{(p)}\equiv\op{\mu}{\nu}_p$ flips the state of spin $p$ to
$\ket\nu$ from $\ket\mu$; and the operator $\Pi_{pq}$ permutes the
states of spins $p$ and $q$.  The permutation operator $\Pi_{pq}$ is
essentially a multilevel generalization of the SU(2)-symmetric spin
interaction $\v s_p\c\v s_q$, with $\v s\equiv\p{\v s_\x,s_\y,s_\z}$,
$s_\alpha\equiv\sigma_\alpha/2$, and $\sigma_\alpha$ a single-spin
Pauli operator.  This permutation operator is equivalently the
multilevel SWAP operation, which merely swaps the state of two spins.

@@

The PS manifold $\M_0$ is spanned by a basis of states
$\ket{m}=\ket{m_1,m_2,\cdots,m_n}$ with a definite occupation number
$m_\mu$ of the single-spin state $\mu$, and $\sum_\mu m_\mu=N$.  We
denote the set of all such assignments of $N$ (identical) spins to $n$
(distinct) states by $\A_n\p{N}$, such that our standard basis for
$\M_0$ is $\set{\ket{m}:m\in\A_n\p{N}}$.  The dimension of $\M_0$ is
then the number of such assignments, which is straightforward to
determine with a standard ``stars and bars'' calculation:
$\abs{\A_n\p{N}} = {N+n-1 \choose n-1} \sim N^{n-1}$.  In the case of
$n=2$, for example, the PS manifold $\M_0$ is precisely the Dicke
manifold of $N$ qubits, spanned by ${N+2-1\choose2-1}=N+1$ Dicke
states $\ket{m_0,m_1}$ for which $m_0+m_1=N$ and the integer $m_0$
($m_1$) indicates the total occupation number of qubit state $\ket{0}$
($\ket{1}$).  The energy of any PS state $\ket\psi\in\M_0$ with
respect to the interaction Hamiltonian $H_0$ in \eqref{eq:H_0} is
\begin{align}
  E_0 = \sum_{\set{p,q}\in\C_N\p{2}} h_{pq}.
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Breaking SU($n$) symmetry}
\label{sec:breaking_sun}

@@

% briefly add expression for the third order Hamiltonian

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Beyond ground-state perturbation theory}
\label{sec:shell_model}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
\label{sec:conclusions}
